---
title: "Josh's Brief R Guide: Basics"
author: Joshua Megnauth
layout: post
output:
  github_document:
    df_print: paged
    toc: true
---
# Introduction

# Vectors
I'd like to keep this short guide practical. So let's start with variables and vectors!

## Defining variables
```{r ages_vector}
ages <- c(29, 24, 10, 42, 65, 8, 56, 22, 15, 8)
```

The `<-` operator _assigns_ the right hand side to the left side. The [c()](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/c) function combines values into a _vector._ Thus, the above code _combines_ 29, 24, 10, 42, 65, 8, 56, 22, 15, and 8 into a _vector_ named _ages._ Let's take a look at our new vector:

```{r ages_vector_print}
ages
```

The vector is the basic unit in R. Vectors hold an arbitrary number of a **single data type**. Keep that fact in mind for the future even if you don't understand what that means as of yet.

R, like other languages, defines a set of basic types. The numbers above default to _numeric_ which are numbers as you'd expect them to be: real numbers. Some of R's other basic types include characters, integers, factors (nominal), logical (booleans), and complex. We'll discuss these types later.

You may define **as many variables** as your little R programmer heart desires. Likewise, R is fairly liberal with variable names.

```{r meow}
meow <- "I love cats and video games!"
```

The restrictions on variable names include: 
* Names are restricted to letters, numbers, periods, and underscores
* Names can't be restricted R keywords, like TRUE or FALSE
* Names must start with a letter, period, or underscore

We can use **meow** and **_meows_and_nyas_** as names but not **99meows**. Likewise, R throws an error if we attempt to name our variable after the keyword _function_.

With that said...you probably shouldn't name your variables _meow._ One ought to name their variables pragmatically and descriptively.

```{r names_examples}
# Decent name - short and descriptive
vg_prices <- c(59, 20, 15, 15, 4.99)

# Bad name - what are somenums? SAS humans do this for some reason.
somenums <- c(295951568, 2, .000000001, 42)

# Bad name - too verbose
selected_pokemon_josh_and_his_best_friend_really_likes_version_two_for_Rlang <- c("Espeon", "Drampa")
```

Naming variables is difficult---shoddy examples aside. Following a canonical method of naming variables will certainly save you headaches later. Trust me.

## Vectors are special

Back to the fun stuff.

Vectors are special data types that are designed to facilitate linear algebra (mostly). Wait! Don't run! Sorry fellow mathemaphobes. I should've warned you ahead of time. We'll scarcely look at any equations here; I only had to mention linear algrebra due to **vectorized operations.**

Statistics and linear algebra are intricately connected. Have you ever calculated a mean? Well, that's linear algebra. Sensibly, the primary data type in R is _totally awesome_ for math.

Vectors may be added to, subtracted from, multiplied by, divided by (et cetera) other vectors. **By the way, a single number, character, et cetera is a vector of one in R.** I'll prove that statement before we continue by using the `==` operator. Two equal signs tests that two statements are equal while `!=` tests for inequality.

```{r vec_of_one}
14 == c(14)
c('j') == 'j'
```

Let's say that the people in our _ages_ vector were exposed to radioactive waste and consequently de-aged by **five.** Let's subtract five from our vector.

```{r age_vec_again}
ages # Take a look at ages again
deaged_ages <- ages - 5 # Subtract 5 and store the result in deaged.ages
deaged_ages # Check out our new vector
```

Wow! **MAGIC.**

You may chain operations together as well. Pretend that Steam is holding a sale where games are 50% off. How would you discount the games in our _vg.prices_ vector?

```{r steam_discount_example}
discount_vg_prices <- vg_prices - vg_prices * .5
discount_vg_prices
```

What happens when an operation is performed on two vectors of unequal size? The smaller vector is reused/recycled per each element of the larger vector---useful and environmentally friendly!

Let's say the radioactive waste alternates between increasing a person's age by five and decreasing the second person's age by five as well.

```{r radioactive_ages_example}
ages + c(5, -5)
```

Take a gander at how the second vector is reused in order to fit the first vector.

# Functions
Functions are reuseable blocks of code. Imagine how you would calculate the mean of the ages vector using what we learned above. You would have to write code to iterate through the vector while summing up each element and storing the sum in a variable. Finally, you would have to divide the sum by the length of the vector.

Here's how the above process would look. Don't worry if you don't understand the code below.

```{r mean_no_functions}
ages_sum <- 0
ages_count <- 0

for (age in ages) {
  ages_sum <- ages_sum + age
  ages_count <- ages_count + 1
}
```

If I had to follow that process every time I had to calculate something as simple as a mean I'd surely change careers.

Functions encapsulate behavior. Calculating a mean is so easy that you surely figured it out already:

```{r ages_mean}
mean(ages)
```

You can view a list of all of base R's functions with the function [builtins()](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/builtins).

## We have to talk about documentation and terminology
R **is not a program.** R is a programming language. Professor Cohen isn't teaching you how to use the R program but how to write R code. The distinction is important because coding is abstract thought much like spoken and written language. Consider this: You've surely taken English classes through your academic life. English professors _do not_ teach how to construct sentences from thoughts. Instead, you read literature and learn how to think about literature and language. Learning R is the same; you're converting your ideas into code.

That line of thought is conceptually important because you cannot view 712 or any other programming, language, or math class as learning by rote. No class will teach everything you need to learn about R. You're learning how to think programatically.

One aspect of thinking programtically is using documentation while coding. You'll sometimes run into problems that reading documentation will help you solve. In generally, you'll likely consult documentation a lot because recalling how every single function works is impossible. I often have several tabs open of docs when coding.

Let's solve a real problem using the docs!

We have two vectors of video game ratings from two different people, A and B (those are their actual names). A and B played different games so some of the observations are missing. You'd like to calculate two means to get a feel for how highly or lowly they rate the selected titles in general.

```{r ratings_mean_broken}
A_vg_ratings <- c(5, 5, 4, NA, 3, 3, 5, 5)
B_vg_ratings <- c(1, 3, NA, NA, 5, 5, 2, 2)

mean(A_vg_ratings)
mean(B_vg_ratings)
```

Woah! What happened? Some base R functions return `NA` if any missing values are present in the data. 

## Reassignment (and why you should avoid it)
R executes a script from top to bottom. Now, some of you may say "well duh...that's obvious!" Yep!

I'll mention yet another "obvious" axiom to continue my tautology as well as to be completely pedantic: lines of code depend on preceding lines of code. In other words, code at the bottom of your scripts depend or may be affected by the code above.

The logic above may seem extreme simple, but I've encountered situations where my peers replaced variables they defined earlier then called functions that expected the variable they replaced!

```{r vector_strnum}
name_and_cats <- c("Josh", 42)
print(name_and_cats)
```
