---
title: "Functions and loading data?"
author: Joshua Megnauth
output:
  github_document:
    df_print: paged
    toc: true
---
# What are functions?

Functions are reusable blocks of code that perform some action. For example, R implements functions for measures of central tendency by default. You can simply call a function rather than calculating the mean or median by verbosely writing code each time. Let's look at how to call a function:

```{r calling_function}
mean(airquality$Wind)
```

The syntax for calling a function is basically like functions in math as well as other programming languages: `name(parameters)`

Functions may take zero or more _parameters_ (also known as _arguments_). Parameters provide input or options to the function. I call the mean function above by passing in the Wind column from the _airquality_ data set.

## We have to talk about parameters!

**No programmer memorizes functions.*

As a programmer you'll likely have documentation open constantly while coding. Let's work through an example of using the documentation and function parameters.

[https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/mean](Open this link please.) Or, alternatively, type `?mean` in your R session (the **Console** tab).

Mean's description states that the function calculates the "(trimmed) arithmetic mean." What does the function mean by "trimmed"? We can see under the _arguments_ section that the function takes a _trim_ parameter which is a "fraction" of "observations to be trimmed from each end of x before the mean is computed." Trimming removes a fraction (such as 0.1 or 10%) of observations before calculating the mean.

Next, let's look at the **na.rm** argument. The parameter **na.rm** "indicat[es] whether NA values should be stripped before the computation proceeds." We can't perform math on `NA` (missing) values.

```{r add_to_na}
14 + NA
```

Thus, the mean function has a parameter to ignore `NA` values to ensure computation actually succeeds.

```{r mean_no_na_rm}
mean(airquality$Ozone)
```

The mean function explicitly returns `NA` because the Ozone column is missing some observations.

```{r mean_na_rm}
mean(airquality$Ozone, na.rm = TRUE)
```

Notice that I called the function with `na.rm = TRUE`. Functions may be called by passing in arguments by position, by name, or mixing the two. The position of each arguments is shown in the documents. Looking back at mean we can see that `x`, our data, is the first parameter. The arguments `trim`, `na.rm`, and variadic follow. I pass `airquality$Ozone` by position which matches the argument `x`. I **have** to use `na.rm = TRUE` because the second argument is `trim`. R wouldn't know what to do if we do something like this: 

```{r mean_by_pos, error=TRUE}
mean(airquality$Ozone, TRUE)
```

The interpreter is trying to match `TRUE` to `trim` as you can see from the error.

## Some useful, built-in functions

Note that this list will clearly not be exhaustive. But I'd like you to at least get an idea of some of the functions available.

* [mean()](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/mean)
* [median()](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/median)
* [lm()](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/lm)
* [glm()](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/glm)
* [dbinom(), pbinom(), qbinom(), rbinom()](https://www.rdocumentation.org/packages/stats/versions/3.3/topics/Binomial)
* [dnorm(), pnorm(), qnorm(), rnorm()](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/Normal)
* [sd](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/sd)
* [read.csv()](https://www.rdocumentation.org/packages/utils/versions/3.6.2/topics/read.table)

# Finally, let's load some data
Type `?read_csv` and `?read.csv`.

The functions `read.csv` and `read_csv` both read CSV files but the latter is a **tidyverse** function that returns tibbles. You can use either for your purposes, but I prefer using tidy functions whenever possible.

Data, like basically everything in computing, is distributed in a passel of different formats which all have their pros and cons---except Excel spreadsheets which are all cons and no pros. CSV stands for comma separated values as you saw in professor Cohen's lecture. CSV's acronym literally tells you what it is: a sequence of values separated by commas. CSVs may be delimited by tabs or semicolons as well. If you looked at the documentation you'd notice that the `delim` parameter can be used to specify a different delimiter.

[An example of a parsed CSV via Neovim](assets/99_functions/csv_example_parsed.png)
[An example of a raw CSV](assets/99_functions/csv_example_raw.png)

# Advanced users
**You may safely ignore this section if you don't care about writing functions.**

(talk about using functions/parameters) and ::
